---
title: "JAGS Tutorial"
output: github_document
---

JAGS is a language that allows you to run Bayesian analyses. It gets at the posterior by generating samples based on the posterior and statistical model.

You'll need to [download and install JAGS](http://mcmc-jags.sourceforge.net/). You can interact with JAGS through one of three R packages:

- `runjags` (recommended for this course)
    - Model written as a single string in R; possibly also allows you to input from file.
    - Quick-start guide vignette: `vignette("quickjags", package="runjags")`
    - Full user guide vignette: `vignette("UserGuide", package="runjags")`
- `rjags` (sample code [here](http://www.johnmyleswhite.com/notebook/2010/08/20/using-jags-in-r-with-the-rjags-package/))
    - Model read in from plain text file.
- `R2jags`
    - this is dependent on `R2winbugs`, which I find doesn't work well outside of Windows machines, so I'm more hesitant to use this package. 
    - Model written in R as a function, but using JAGS language; or inputted from file.

Also, the `coda` package is useful for working with the output of at least `runjags`. 

About the JAGS language:

- Generates samples of parameters based on the prior and statistical model.
- Need to specify which parameters you want to include in the output (aka "track" or "monitor").
- Specify probability distributions similarly to R, except:
    - Draw samples using calls like `dexp` and `dnorm`, not `rexp` and `rnorm`. 
    - The JAGS version of `rnorm` uses the precision (=1/variance) instead of standard deviation. 
The documentation of JAGS code is not as nice as R. You have to look things up from a table-of-contents-style search from [this](http://www.stats.ox.ac.uk/~nicholls/MScMCMC15/jags_user_manual.pdf) document. Page 29 shows the aliases for various distributions. 

For this week's lab assignment (3), you'll only be using it to generate observations from a distribution. Let's generate data from a N(0,2) distribution (that is, variance=2), and ignore the warning messages for this week. 

```{r}
suppressPackageStartupMessages(library(runjags))
my_model <- "
model{
    # This is a comment.
    theta ~ dnorm(0, 1/2)
}
"
fit <- run.jags(my_model, monitor="theta", n.chains=1, sample=1000)
theta <- coda::as.mcmc(fit)
head(theta)
plot(theta)
```


## More sample code

```{r}
library(tidyverse)
library(runjags)
n <- 50
dat <- tibble(x=rnorm(n),
              y=x + rnorm(n))
jagsdat <- c(as.list(dat), n=nrow(dat))
model <- "model{
    for (i in 1:n) {
        y[i] ~ dnorm(beta*x[i], tau)
    }
    tau <- pow(sigma, -2)
	sigma ~ dunif(0, 100)
    beta ~ dnorm(0, 0.001)
}"
foo <- run.jags(
    model=model, 
    monitor=c("beta0", "beta1", "sigma"), 
    data=jagsdat
)
plot(coda::as.mcmc(foo))
```

